---
title: "Georgia Election Project"
author: "Sejal Govindarao"
date: "`r format(Sys.time(), '%B %d, %Y')`"
format:
  html:
    self-contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#Load libraries 
library(tidyverse) 
library(janitor) 
library(readxl) 
library(scales)
library(lubridate)
library(tidyverse)
library(tigris)
library(sf)
library(tmap)
library(tmaptools)
library(htmltools)
library(janitor)
library(rmapshaper)
library(here)
options(tigris_class = "sf")

```

# Introduction

In this analysis walkthrough, we will be looking at Georgia election results by county for U.S. Senate and Governor in 2022 and President in 2020.

```{r}
#import the GA election data for this assignment

#GA senate
senate_results <- read_excel(here("ga_votes.xlsx"), sheet = "senate")

#GA governor
governor_results <- read_excel(here("ga_votes.xlsx"), sheet = "governor")


```

Two columns in particular to highlight that we'll use to examine how the Republican Senate and Governor candidates did compared with former President Trump's performance in 2020:

-   **`pct_r_above_trump`**. This column calculates the difference in percentage points between what the 2022 Republican candidate got (`pct_r_2022`) and what Trump got two years earlier (`pct_r_2020`). A negative value means the candidate did worse than Trump, a positive value means they did better.

-   **`is_r_above_trump`**. This companion columns is a so-called "flag" column, which provides a Yes/No as to whether the Republican candidate did better or worse than Trump. In other words, is the `pct_r_above_trump` value positive or negative.

# Charts

First, we will examine the GA Senate race in 2022 This bar chart compares the counties in which Republican candidate Hershel Walker did better vs. worse than Trump.

```{r}

SenateRepublicanCandidateAsComparedToTrump <- senate_results %>% 
count(is_r_above_trump)

ggplot(SenateRepublicanCandidateAsComparedToTrump, aes(x = is_r_above_trump, y = n)) +
  geom_col()



```


Next, we will do the same for the GA Gubernatorial race in 2022. This bar chart compares the counties in which Republican candidate Brian Kemp did better vs. worse than Trump.

```{r}

RGovsComparedToTrump <- governor_results %>% 
count(is_r_above_trump)

ggplot(RGovsComparedToTrump, aes(x = is_r_above_trump, y = n)) +
  geom_col()



```



Now, again examining the 2022 GA Senate race, this bar chart illustrates the top 10 counties with the biggest drops between Walker and Trump.

```{r}
Top10 <- senate_results %>% 
  arrange(pct_r_above_trump) %>% 
  head(10)

ggplot(Top10, aes(x = reorder(county,pct_r_above_trump), y = pct_r_above_trump)) +
  geom_col() +
  coord_flip()


```



This looks at the top 10 counties in the Senate Race with highest victory margins for Raphael Warnock.

```{r}

MarginOfVictory <- senate_results %>% 
  mutate(margin_vic = (pct_d_2022 - pct_r_2022)) 

Senate10MOV <- MarginOfVictory %>% 
  arrange(desc(margin_vic)) %>% head(10)

ggplot(Senate10MOV, aes(x = reorder(county, -margin_vic), y = margin_vic)) + 
  geom_col()



```

This looks at the top 10 counties in the Gubernatorial Race with highest victory margins for Stacey Abrams.

```{r}

MOV_GOV <- governor_results %>% 
  mutate(govMOV = (pct_d_2022 - pct_r_2022))

Gov10MOV <- MOV_GOV %>% 
  arrange(desc(govMOV)) %>%  head(10)

ggplot(Gov10MOV, aes(x = reorder(county, -govMOV), y = govMOV)) + 
  geom_col()



```

# Maps



```{r}
counties_geo <- tigris::counties(state = "GA", resolution = "20m", cb = TRUE)
```

Now, we will visualize Georgia counties during the 2022 election!

```{r}

tm_shape(counties_geo) + 
  tm_polygons()


```

Here we see a map of Georgia counties where Walker did better vs. worse than Trump.

```{r}

senatemap <- inner_join(counties_geo, senate_results, by = c("GEOID" = "fipsCode"))
tm_shape(senatemap) +
  tm_polygons("is_r_above_trump", id = "GEOID")



```

Switching over to the Gubernatorial race, here we see a map of Georgia counties where Kemp did better vs. worse than Trump.

```{r}

govmap <- inner_join(counties_geo, governor_results, by = c("GEOID" = "fipsCode"))
tmap_mode(mode = "plot")
tm_shape(govmap) +
  tm_polygons("is_r_above_trump", id = "GEOID")



```

This map shows us Raphael Warnock's victory margin per county. 


```{r}

SenVictoryMargins <- inner_join(counties_geo, MarginOfVictory, by = c("GEOID" = "fipsCode"))
tmap_mode(mode = "plot")

tm_shape(SenVictoryMargins) + 
  tm_polygons("margin_vic", id = "GEOID")



```

This map shows us Stacey Abrams' victory margin per county. 

```{r}

GovChloroMap <- inner_join(counties_geo, MOV_GOV, by = c("GEOID" = "fipsCode"))
tmap_mode(mode = "plot")

tm_shape(GovChloroMap) + 
  tm_polygons("govMOV", id = "GEOID")



```

# 
