---
title: "Virginia Election Project"
author: "Sejal Govindarao"
execute:
  echo: true
format:
  html:
    self-contained: true
    code-tools: true
---


```{r setup, include=FALSE, warning=FALSE, message=FALSE}


library(tidyverse)
library(janitor)
library(httr)
library(jsonlite)
library(kableExtra)
library(here)
options(scipen = 999)
options(stringsAsFactors = FALSE)


#load saved joined data file from previous script
joined_vacomparison <- readRDS(here("processed_data", "joined_vacomparison.rds"))



```


# Comparing Virginia Election Results by County
In this analysis walkthrough, we will be looking at Virginia election results by county. We will compare the 2020 Presidential Election and 2021 Gubernatorial Election -- this is particularly interesting because Biden, a Democrat, won in 2020 while Youngkin, a Republican, won in 2021. 

In this first table, we will see a breakdown by county of the margins by which Trump, Biden, McAuliffe and Youngkin faired. Check out the interactive data table!


```{r}

DT::datatable(joined_vacomparison)


```

This next table adds a column at the end, illustrating how many points Youngkin held over Trump in each respective county.

```{r}

joined_vacomparison = joined_vacomparison %>% 
  mutate(YoungkinADV = pct_youngkin - trump_pct)
head(joined_vacomparison)



```
Next, we have a bar chart that hones in on the analysis we saw in the prior table -- this chart focuses on the top 5 counties where Youngkin in 2021 held the highest percentage of the vote as compared to Trump in 2020.  

```{r}

TOP5 <- joined_vacomparison %>% 
  arrange(desc(YoungkinADV)) %>% 
  slice(1:5)

ggplot(TOP5, aes(x = locality, y = YoungkinADV)) +
  geom_bar(stat="identity", fill="coral")+
  scale_x_discrete(guide = guide_axis(n.dodge=3))

```

This bar chart isolates Youngkin in 2021, illustrating the top 5 counties he performed best in.

```{r}

YOUNGKIN_5 <- joined_vacomparison %>% 
  arrange(desc(pct_youngkin)) %>% 
  slice(1:5)

ggplot(YOUNGKIN_5, aes(x = locality, y = pct_youngkin)) +
  geom_bar(stat="identity", fill="darkred")+
  scale_x_discrete(guide = guide_axis(n.dodge=3))
```



Similarly, this bar chart illustrates the top 5 counties McAuliffe performed best in during the 2021 VA Election. 

```{r}

MCAULIFFE_5 <- joined_vacomparison %>% 
  arrange(desc(pct_mcauliffe)) %>% 
  slice(1:5)

ggplot(MCAULIFFE_5, aes(x = locality, y = pct_mcauliffe)) +
  geom_bar(stat="identity", fill="steelblue")+
  scale_x_discrete(guide = guide_axis(n.dodge=3))

```


Here we see a comparison between Biden and McAuliffe, illustrating the top 5 counties Biden outperformed McAuliffe in. 

```{r}

joined_vacomparison = joined_vacomparison %>% 
  mutate(biden_adv = biden_pct - pct_mcauliffe)

BIDEN_ADV <- joined_vacomparison %>% 
  arrange(desc(biden_adv)) %>% 
  slice(1:5)

ggplot(BIDEN_ADV, aes(x = locality, y = biden_adv)) +
  geom_bar(stat="identity", fill="navy")+
  scale_x_discrete(guide = guide_axis(n.dodge=3))

```


We now see the top 5 counties in Virginia Biden outperformed Trump in 2020. 

```{r}

joined_vacomparison = joined_vacomparison %>% 
  mutate(biden20_adv = biden_pct - trump_pct)

BIDEN20_ADV <- joined_vacomparison %>% 
  arrange(desc(biden20_adv)) %>% 
  slice(1:5)

ggplot(BIDEN20_ADV, aes(x = locality, y = biden20_adv)) +
  geom_bar(stat="identity", fill="purple")+
  scale_x_discrete(guide = guide_axis(n.dodge=3))

```

And, conversely, here are the top 5 counties in Virginia Trump outperformed Biden in 2020. 

```{r}

joined_vacomparison = joined_vacomparison %>% 
  mutate(trump20_adv = trump_pct - biden_pct)

TRUMP20_ADV <- joined_vacomparison %>% 
  arrange(desc(trump20_adv)) %>% 
  slice(1:5)

ggplot(TRUMP20_ADV, aes(x = locality, y = trump20_adv)) +
  geom_bar(stat="identity", fill="orange")+
  scale_x_discrete(guide = guide_axis(n.dodge=3))

```





